---
title: "Final Project: Decoding Diabetes – Analyzing Key Health Indicators"
author: "Jothish Kumar Polaki"
date: "2024-12-13"
output: html_document
runtime: shiny
---


## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r importing and loading libraries, warning=FALSE, message=FALSE}
if(!require(pacman)) #<1>
  install.packages("pacman")

pacman::p_load(tidyverse,
               ggplot2,
               readr,
               dplyr,
               reshape2,
               caret,
               pROC,
               randomForest,
               glmnet,
               shiny
               )

```

```{r loading data, warning=FALSE, message=FALSE}
data <- read_csv("diabetes_binary_5050split_health_indicators_BRFSS2015.csv")

# checking top rows
head(data)

# checking for NA's in data
colSums(is.na(data))

```

## EDA Pre-processing

-   Changing the values in the selected columns for better understanding.
```{r preprocessing, warning=FALSE, message=FALSE}
# getting required data for analysis
eda_data <- data |>
  select(
    Diabetes_binary, AnyHealthcare, NoDocbcCost, HighBP, HighChol, 
    Income, Education, Age, Sex, BMI
  ) |>
  mutate(
    # changing binary values to "No" and "Yes"
    AnyHealthcare = ifelse(AnyHealthcare == 1, "Yes", "No"),
    NoDocbcCost = ifelse(NoDocbcCost == 1, "Yes", "No"),
    HighBP = ifelse(HighBP == 1, "Yes", "No"),
    HighChol = ifelse(HighChol == 1, "Yes", "No"),
    
    # updating the income values to meaning labels
    Income = case_when(
      Income == 1 ~ "Less than $10,000",
      Income == 2 ~ "Less than $15,000",
      Income == 3 ~ "Less than $20,000",
      Income == 4 ~ "Less than $25,000",
      Income == 5 ~ "Less than $35,000",
      Income == 6 ~ "Less than $50,000",
      Income == 7 ~ "Less than $75,000",
      Income == 8 ~ "$75,000 or more",
      TRUE ~ as.character(Income)  # Keep original value if outside 1-8
    ),
    # updating education levels
    Education = case_when(
      Education == 1 ~ "Never attended school or only kindergarten",
      Education == 2 ~ "Grades 1 through 8 (Elementary)",
      Education == 3 ~ "Grades 9 through 11 (Some high school)",
      Education == 4 ~ "Grade 12 or GED (High school graduate)",
      Education == 5 ~ "College 1 year to 3 years (Some college or technical school)",
      Education == 6 ~ "College 4 years or more (College graduate)",
      TRUE ~ as.character(Education)  # Keep original value if outside 1-6
    ),
    # updating the age column
    Age = case_when(
      Age == 1 ~ "18-24",
      Age == 2 ~ "25-29",
      Age == 3 ~ "30-34",
      Age == 4 ~ "35-39",
      Age == 5 ~ "40-44",
      Age == 6 ~ "45-49",
      Age == 7 ~ "50-54",
      Age == 8 ~ "55-59",
      Age == 9 ~ "60-64",
      Age == 10 ~ "65-69",
      Age == 11 ~ "70-74",
      Age == 12 ~ "75-79",
      Age == 13 ~ "80 or older",
      TRUE ~ as.character(Age)  # Keep original value if outside 1-13
    )
  )

#view(eda_data)
```


## Exploratory Data Analysis

### Question 1:

-   What is the relationship between socioeconomic factors like Income, Education and Healthcare access and the prevalance of prediabetes or diabetes?

Education Distributed by Diabetes Status:

```{r education, warning=FALSE, message=FALSE}
education_data <- eda_data |>
  select(Education, Diabetes_binary)

education_data <- education_data %>%
  mutate(EducationNumeric = case_when(
    Education == "Never attended school or only kindergarten" ~ 1,
    Education == "Grades 1 through 8 (Elementary)" ~ 2,
    Education == "Grades 9 through 11 (Some high school)" ~ 3,
    Education == "Grade 12 or GED (High school graduate)" ~ 4,
    Education == "College 1 year to 3 years (Some college or technical school)" ~ 5,
    Education == "College 4 years or more (College graduate)" ~ 6
  ))

# education vs diabete plot
education_data |>
  ggplot(aes(x = EducationNumeric, fill = as.factor(Diabetes_binary))) +
  geom_histogram(position = "dodge", binwidth = 0.5, color = "black") +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"),
                    name = "Diabetes Status",
                    labels = c("No Diabetes", "Yes Diabetes")) +
  scale_x_continuous(breaks = 1:6, labels = c(
    "Never attended or Kindergarten",
    "Grades 1-8 (Elementary)",
    "Grades 9-11 (Some HS)",
    "Grade 12 or GED (HS Graduate)",
    "College 1-3 years",
    "College 4+ years"
  )) +
  labs(
    x = "Education levels",
    y = "Count",
    title = "Distribution of Education Levels by Diabetes Status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7), legend.position = "top")
```

Income Distribution by Diabetes status:

```{r income, warning=FALSE, message=FALSE}
income_data <- eda_data |>
  select(Income, Diabetes_binary)

income_data <- income_data |>
  mutate(
    incomeNumeric = case_when(
      Income == "$75,000 or more" ~ 8,
      Income == "Less than $75,000" ~ 7,
      Income == "Less than $50,000" ~ 6,
      Income == "Less than $35,000" ~ 5,
      Income == "Less than $25,000" ~ 4,
      Income == "Less than $20,000" ~ 3,
      Income == "Less than $15,000" ~ 2,
      Income == "Less than $10,000" ~ 1,
      TRUE ~ NA_real_
  )
)

# income vs diabetes plot
income_data |>
  ggplot(aes(x = incomeNumeric, fill = factor(Diabetes_binary))) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"),
                    name = "Diabetes Status",
                    labels = c("No Diabetes", "Yes Diabetes")) +
  geom_histogram(position = "dodge", binwidth = 0.5, color = "black") +
  scale_x_continuous(breaks = 1:8, labels = c(
    "Less than $10,000",
    "Less than $15,000",                                      "Less than $20,000",                                      "Less than $25,000",                                      "Less than $35,000",                                      "Less than $50,000",                                      "Less than $75,000",                                      "$75,000 or more"
  )) +
  labs(
    x = "Income Levels",
    y = "Count",
    fill = "Diabetes Status",
    title = "Income Distribution by Diabetes Status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top")
```

Healthcare Access:

```{r healthcare, warning=FALSE, message=FALSE}
# any healthcare coverage
eda_data |>
  ggplot(aes(x = factor(AnyHealthcare), fill = factor(Diabetes_binary))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"),
                    name = "Diabetes Status",
                    labels = c("No Diabetes", "Yes Diabetes")) +
  labs(
    x = "Has any healthcare coverage",
    y = "Proportion",
    fill = "Diabetes status",
    title = "Distribution of Healthcare Access by diabetes status"
  ) +
  theme_minimal()

# Unable to see a doctor due to financial issues
eda_data |>
  ggplot(aes(x = factor(NoDocbcCost), fill = factor(Diabetes_binary))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"),
                    name = "Diabetes Status",
                    labels = c("No Diabetes", "Yes Diabetes")) +
  labs(
    x = "Could not see a doctor due to financial constraints",
    y = "Proportion",
    fill = "Diabetes status",
    title = "Financial barrier to Healthcare Access by diabetes status"
  ) +
  theme_minimal()

```

### Question 2:

-   How do comorbid conditions like high blood pressure, high cholesterol and heart diseases relate to the risk of prediabetes or diabetes?

Distribution of high BP and cholesterol by Diabetes status

```{r comorbid, warning=FALSE, message=FALSE}
# proportion of high BP by diabetes status
eda_data |>
  ggplot(aes(x = factor(HighBP), fill = factor(Diabetes_binary))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"),
                    name = "Diabetes Status",
                    labels = c("No Diabetes", "Yes Diabetes")) +
  labs(
    x = "High Blood Pressure",
    y = "Proportion",
    fill = "Diabetes status",
    title = "Prevelance of High BP by diabetes status"
  ) +
  theme_minimal()

# proportion of high cholesterol by diabetes status
eda_data |>
  ggplot(aes(x = factor(HighChol), fill = factor(Diabetes_binary))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"),
                    name = "Diabetes Status",
                    labels = c("No Diabetes", "Yes Diabetes")) +
  labs(
    x = "High Cholesterol",
    y = "Proportion",
    fill = "Diabetes status",
    title = "Prevelance of High Cholesterol by diabetes status"
  ) +
  theme_minimal()

```

Heatmap to visualize comorbidites:

```{r heatmap, warning=FALSE, message=FALSE}
comorbidity_data <- data |>
  select(Diabetes_binary, HighBP, HighChol, Stroke, HeartDiseaseorAttack) |>
  group_by(Diabetes_binary) |>
  summarise_all(mean)

melted_data <- melt(comorbidity_data, id.vars = "Diabetes_binary")

# heatmap of comorbid conditions vs diabetes
melted_data |>
  ggplot(aes(x = variable, y = factor(Diabetes_binary), fill = value)) +
  geom_tile() +
  labs(
    x = "Comorbidites",
    y = "Diabetes Status",
    fill = "Proportion",
    title = "Heatmap of Comorbidites by Diabetes status"
  ) +
  scale_y_discrete(labels = c("0" = "No", "1" = "Yes")) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12))

```

Analyzing confounders like **Age**, **Sex** and **BMI**:

```{r confounder, warning=FALSE, message=FALSE}
confounder_data <- eda_data |>
  mutate(BMI_Category = cut(BMI, breaks = c(0, 18.5, 25, 30, Inf), 
                            labels = c("Underweight", "Normal", "Overweight", "Obese")),
         Sex = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
         Diabetes_binary = factor(Diabetes_binary, levels = c(0, 1), labels = c("No Diabetes", "Diabetes")))

# Faceted bar plot
confounder_data |>
  ggplot(aes(x = factor(Age), fill = BMI_Category)) +
  geom_bar(position = "fill") +
  facet_wrap(Sex ~ Diabetes_binary) +
  labs(x = "Age", y = "Proportion", fill = "BMI Category") +
  ggtitle("Age, Sex, and BMI Distribution by Diabetes Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top")
```

## Machine learning model

### Logistic Regression

```{r ml pre processing, warning=FALSE, message=FALSE}
# selecting predictors and the outcome
ml_data <- eda_data |>
  select(Diabetes_binary, AnyHealthcare, NoDocbcCost, Income, Education, HighBP, HighChol, BMI, Age, Sex)

# converting categorical variables to factors
ml_data <- ml_data |>
  mutate(
    Diabetes_binary = as.factor(Diabetes_binary),
    AnyHealthcare = as.factor(AnyHealthcare),
    NoDocbcCost = as.factor(NoDocbcCost),
    Income = as.factor(Income),
    Education = as.factor(Education),
    HighBP = as.factor(HighBP),
    HighChol = as.factor(HighChol),
    Age = as.factor(Age),
    Sex = as.factor(Sex)
  )

# splitting the data to train and test set
set.seed(123)
train_index <- createDataPartition(ml_data$Diabetes_binary, p = 0.8, list = FALSE)
train_data <- ml_data[train_index, ]
test_data <- ml_data[-train_index, ]

dim(train_data)
dim(test_data)
```


```{r lopmodel, warning=FALSE, message=FALSE}
# creating a log reg model
log_model <- glm(Diabetes_binary ~ ., data = train_data, family = binomial)

#summary(log_model)

# predicting probabilities
test_predictions <- predict(log_model, test_data, type = "response")
# converting probablities into binary
test_data$predicted <- ifelse(test_predictions > 0.5, "1", "0")

# generating a confusion matrix
conf_matrix <- confusionMatrix(as.factor(test_data$predicted), test_data$Diabetes_binary, positive = "1")

print(conf_matrix)

```


```{r roc, warning=FALSE, message=FALSE}
# plotting a roc curve
roc_curve <- roc(test_data$Diabetes_binary, as.numeric(test_predictions))

plot(roc_curve, col = "blue", main = "ROC Curve for Logistic Regression")
```

### Random Forest

```{r rf, warning=FALSE, message=FALSE}

set.seed(123)
# creating a random forest model
rf_model <- randomForest(
  Diabetes_binary ~ ., 
  data = train_data, 
  ntree = 500, 
  mtry = 3, 
  importance = TRUE
)

print(rf_model)

# getting predictions
rf_predictions <- predict(rf_model, test_data)

# making a confusion matrix from the model
conf_matrix_rf <- confusionMatrix(rf_predictions, test_data$Diabetes_binary, positive = "1")
print(conf_matrix_rf)

# plotting a feature importance plot
varImpPlot(rf_model, main = "Feature Importance")

```

## Shiny App

### UI

```{r ui, warning=FALSE, message=FALSE}
# creating an UI for the app
ui <- fluidPage(
  titlePanel("Diabetes Risk Explorer"),
  sidebarLayout(
    sidebarPanel(
      selectInput("eda_var", "Select Variable for EDA:", 
                  choices = c("Education", "Income", "Age", "Sex", "BMI", "Income", "HighBP", "HighChol"), selected = "BMI"),
      selectInput("eda_view", "Choose EDA View:", 
                  choices = c("Distribution", "Heatmap", "Proportions"), selected = "Distribution"),
      numericInput("input_age", "Enter Age:", value = 40, min = 18, max = 100),
      numericInput("input_bmi", "Enter BMI:", value = 25, min = 10, max = 50),
      checkboxInput("input_highbp", "Has High Blood Pressure?", value = FALSE),
      checkboxInput("input_highchol", "Has High Cholesterol?", value = FALSE),
      selectInput("input_healthcare", "Do you have healthcare coverage?", 
                  choices = c("Yes", "No")),
      selectInput("input_nodoc", "Could not see a doctor due to cost?", 
                  choices = c("Yes", "No")),
      selectInput("input_income", "Select Income level:", 
                  choices = c("Less than $10,000", "Less than $15,000", 
                              "Less than $20,000", "Less than $25,000", 
                              "Less than $35,000", "Less than $50,000", 
                              "Less than $75,000", "$75,000 or more")),
      selectInput("input_education", "Select Education level:", 
                  choices = c("Never attended school or only kindergarten", 
                              "Grades 1 through 8 (Elementary)", 
                              "Grades 9 through 11 (Some high school)", 
                              "Grade 12 or GED (High school graduate)", 
                              "College 1 year to 3 years (Some college or technical school)", 
                              "College 4 years or more (College graduate)")),
      radioButtons(
  inputId = "input_sex",
  label = "Select Sex:",
  choices = c("Male", "Female"),
  selected = NULL
),
      actionButton("predict", "Predict Diabetes Risk")
    ),
    mainPanel(
      tabsetPanel(
        tabPanel("EDA", plotOutput("eda_plot")),
        tabPanel("Prediction", verbatimTextOutput("pred_output")),
        tabPanel("Feature Importance", plotOutput("feature_plot")),
        tabPanel("Data Summary", verbatimTextOutput("data_summary"))
      )
    )
  )
)

```

### Server

```{r server, warning=FALSE, message=FALSE}
# creating server to run the app
server <- function(input, output) {
  
  # EDA plots
  output$eda_plot <- renderPlot({
    if (input$eda_var %in% c("HighBP", "HighChol", "AnyHealthcare", "NoDocbcCost")) {
      # For categorical variables
      ggplot(eda_data, aes_string(x = input$eda_var, fill = "factor(Diabetes_binary)")) +
        geom_bar(position = "fill") +
        scale_fill_manual(
          values = c("0" = "red", "1" = "blue"),
          name = "Diabetes Status",
          labels = c("No Diabetes", "Yes Diabetes")
        ) +
        labs(
          x = input$eda_var,
          y = "Proportion",
          fill = "Diabetes Status",
          title = paste("Proportion of", input$eda_var, "by Diabetes Status")
        ) +
        theme_minimal()
    } else {
      # For continuous or numeric variables like Income or Age
      ggplot(eda_data, aes_string(x = input$eda_var, fill = "factor(Diabetes_binary)")) +
        geom_histogram(
          position = "dodge", 
          binwidth = 1, 
          color = "black"
        ) +
        scale_fill_manual(
          values = c("0" = "red", "1" = "blue"),
          name = "Diabetes Status",
          labels = c("No Diabetes", "Yes Diabetes")
        ) +
        labs(
          x = input$eda_var,
          y = "Count",
          fill = "Diabetes Status",
          title = paste("Distribution of", input$eda_var, "by Diabetes Status")
        ) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    }
  })
  
  # predictions
  observeEvent(input$predict, {
  # preparing data for predictions
  user_data <- data.frame(
    AnyHealthcare = factor(input$input_healthcare, levels = levels(ml_data$AnyHealthcare)),
    NoDocbcCost = factor(input$input_nodoc, levels = levels(ml_data$NoDocbcCost)),
    HighBP = factor(ifelse(input$input_highbp, "Yes", "No"), levels = levels(ml_data$HighBP)),
    HighChol = factor(ifelse(input$input_highchol, "Yes", "No"), levels = levels(ml_data$HighChol)),
    BMI = as.numeric(input$input_bmi),
    Age = factor(case_when(
      input$input_age >= 18 & input$input_age <= 24 ~ "18-24",
      input$input_age >= 25 & input$input_age <= 29 ~ "25-29",
      input$input_age >= 30 & input$input_age <= 34 ~ "30-34",
      input$input_age >= 35 & input$input_age <= 39 ~ "35-39",
      input$input_age >= 40 & input$input_age <= 44 ~ "40-44",
      input$input_age >= 45 & input$input_age <= 49 ~ "45-49",
      input$input_age >= 50 & input$input_age <= 54 ~ "50-54",
      input$input_age >= 55 & input$input_age <= 59 ~ "55-59",
      input$input_age >= 60 & input$input_age <= 64 ~ "60-64",
      input$input_age >= 65 & input$input_age <= 69 ~ "65-69",
      input$input_age >= 70 & input$input_age <= 74 ~ "70-74",
      input$input_age >= 75 & input$input_age <= 79 ~ "75-79",
      input$input_age >= 80 ~ "80 or older"
    ), levels = levels(ml_data$Age)),
    Income = factor(input$input_income, levels = levels(ml_data$Income)),
    Education = factor(input$input_education, levels = levels(ml_data$Education)),
    Sex = factor(
  ifelse(input$input_sex == "Male", "0", "1"),
  levels = levels(ml_data$Sex)
)
  )
  user_data$predicted_prob <- predict(log_model, newdata = user_data, type = "response")
  output$pred_output <- renderText({
    paste("Predicted Probability of Diabetes: ", round(user_data$predicted_prob * 100, 2), "%")
  })
})
  
  # Feature Importance
  output$feature_plot <- renderPlot({
    varImpPlot(rf_model)  # Random forest feature importance plot
  })
  
  # Data Summary
  output$data_summary <- renderPrint({
    summary(eda_data)
  })
}


```

Running the app 

```{r shinyapp, warning=FALSE, message=FALSE}
shinyApp(ui = ui, server = server)

```